A Bayesian model of latitudinal temperature gradients applied to the Early Eocene

### 2023-02-23 manuscript/manuscript.Rmd
Update manuscript.Rmd

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`) degC. The inferred latitudinal gradient is somewhat steeper (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC) in the northern hemisphere, but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig. S3).)

@@

## Early Eocene climate
The geochemical proxy record and ecological data agree that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data. The modeled SST of @Tierney2017 using TEX86 is ~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator. @Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC. All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC. @Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere.
Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.

@@

GMSsT

>>>>

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`) degC. The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig. S3).)

@@

## Early Eocene climate
The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data. The modeled SST of @Tierney2017 using TEX86 is ~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator. @Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC. All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC. @Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere.
Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.

@@

Our GMSsT estimate is
